---
title: "Preliminary COVID-19 models"
author: "David Nguyen"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

# Sensitivity analysis
I am varying the proportional contact rate under self-isolation and social distancing.

* Individuals with severe symptoms have 0 contacts.
* Individuals with mild symptoms have 10 - 30 % contacts compared to normal.
* Inviduals under social distancing have 10 - 30 % of contacts compared to normal.

I also assume that $\beta_\text{isolated, mild symptoms} \leq \beta_\text{social dist}$

Going down the page, the plots show scenarios where distancing and isolation are less effective.

```{r simulation, cache = TRUE, include = FALSE}
## note, need to figure out how to invalidate cache when I change source("model code") 
# define params
covid_params <- c(
  beta0            = 0.5
  , Ca               = 1
  , Cp               = 1
  , Cs               = 1
  , Cm               = 1
  , alpha            = 1/3
  , gamma            = 1/5.2
  , lambda_a         = 1/7
  , lambda_s         = 1/4
  , lambda_m         = 1/7
  , lambda_p         = 1/0.5
  , rho              = 1/10.7
  , delta            = 0.2
  , mu               = 19/20
  , N                = 5e6 
  , E0               = 20
)

# time line
years <- 5
sim_length <- round(365.25*years)

# make matrix of policy choices
control_pars <- 
  expand.grid(## policy choices
              t_nothing = 60             # days at beginning of outbreak w/o control
              ,t_social_dist = c(30, 60, 90) # days of social distancing
              ,control_policy = c(0,1,2)
              ## intervention effects
              ,c_iso_severe = 0
              ,c_iso_mild   = c(0.1, 0.2, 0.3)
              ,c_social_dist = c(0.1, 0.2, 0.3)
              #,c_thresh_int = c(0.2, 0.3, 0.4)
              # policy rules
              ,start_thresh_int = 20
              ,stop_thresh_int  = 2
              )

# adjust control pars
# assume that beta reduction under soc dist == surfing (when control on)
control_pars$c_thresh_int <- control_pars$c_social_dist 
# assume that contacts of iso_mild <= other social dist
control_pars <- control_pars[control_pars$c_iso_mild <= control_pars$c_social_dist,] 

# init simulation
sim_list <- vector("list", length = nrow(control_pars))
n_sim <- 50

# run simulations
for (i in seq_along(sim_list)) {
  
  # set policy choices
  t_nothing <- control_pars[i, "t_nothing"]
  t_social_dist <- control_pars[i, "t_social_dist"]
  control_policy <- control_pars[i, "control_policy"]
  
  # effect of interventions (proportion reduction of beta)
  c_iso_severe <- control_pars[i, "c_iso_severe"]
  c_iso_mild   <- control_pars[i, "c_iso_mild"]
  c_social_dist <- control_pars[i, "c_social_dist"]
  c_thresh_int <- control_pars[i, "c_thresh_int"]
  
  # surfing rules
  start_thresh_int <- control_pars[i, "start_thresh_int"] # number of hospitalized cases
  stop_thresh_int <- control_pars[i, "stop_thresh_int"]   # number of hostpitalized cases
  
  source("mk_covid-19_setup.R")
  
  # create df of sim pars
  sim_pars <- replicate(n_sim * sim_length, control_pars[i,], simplify = FALSE) %>% bind_rows()
  
  # simulate with a set of parameters
  sim_list[[i]] <- do.call(
    pomp::simulate
    , list(
      object       = covid
      , params       = covid_params
      , nsim         = n_sim
      , format       = "d"
      , include.data = F
      , seed         = 1001)
  ) %>% cbind(sim_pars)
  
}

sim_df <- bind_rows(sim_list)
rm(sim_list)
# rename control policies
sim_df <-
  sim_df %>%
  mutate(control_policy = case_when(control_policy == 0 ~ "end social distancing",
                                    control_policy == 1 ~ "continue social distancing",
                                    control_policy == 2 ~ "use surfing policy"))
```

# time series

```{r make_ts_plots}
# plot total infected and hospitalized
sample_ids <- base::sample(seq_along(1:n_sim), 10)

reduction_iso_mild <- unique(control_pars$c_iso_mild)
reduction_intervention <- unique(control_pars$c_social_dist)

ts_plot <- vector("list", length = 6)
plot_no <- 0
for(i in seq_along(reduction_intervention)) {
  for(j in seq_along(reduction_iso_mild)) {
    if (reduction_iso_mild[j] > reduction_intervention[i]) next
    plot_no <- plot_no + 1
    ts_plot[[plot_no]] <-
      sim_df  %>%
      filter(.id %in% sample_ids,
             c_social_dist == reduction_intervention[i],
             c_iso_mild == reduction_iso_mild[j]) %>%
      ggplot() +
      geom_line(aes(x = day, y = Ia + Ip + Is, group = .id), col = "purple", alpha = 0.5) +
      geom_line(aes(x = day, y = H, group = .id), col = "darkgreen", alpha = 0.5) +
      geom_point(aes(x = day, y = 0, col = as.factor(thresh_crossed)), 
                 shape = 22, size = 1, alpha = 0.5) + # show surf policy at bottom
      scale_color_manual(values = c(NA, "black")) +
      scale_y_log10() +
      facet_grid(t_social_dist ~ control_policy) +
      labs(title = "Epidemic scenarios under alternative management strategies",
           subtitle = paste("Contact reduction for mild and isolated: ", reduction_iso_mild[j], "\n",
                            "Contact reduction during intervention: ", reduction_intervention[i], sep = ""),
           y = "(log10) number of people",
           x = "days",
           caption = "purple is total infected and non-hospitalized, green is hospitalized cases") +
      theme(legend.position = "n")
    
  }
}
```

Note that stochastic fade-outs under surfing policies are more common when contacts are strongly reduced and when the initial social distancing order lasts for longer periods of time.

```{r plot_ts, warning = FALSE, eval = TRUE}
ts_plot[[1]]  
ts_plot[[2]]  
ts_plot[[3]]  
ts_plot[[4]]  
ts_plot[[5]]  
ts_plot[[6]]  
```

# Efficient frontier

I think the variability in the person-days under shut down for the surfing policy are caused by the stochastic fade outs. This variability shrinks as behavioral changes are less effective at reducing contacts since fade-outs become less common.

Note that the x-axis is on the log10 scale here, so the differences between surfing and continuous social distancing are quite larger than they first appear.

```{r}
summary_df <-
  sim_df %>%
  group_by(.id, t_social_dist, control_policy, c_social_dist, c_iso_mild) %>%
  # mutate() %>%
  mutate(total_death = last(D, order_by = day),                                      # get cumulative death over simulation
         pd_surf = ifelse(policy == 2, (thresh_crossed*(S + E + Ia + Ip + R)), 0),   # get daily total ppl under control
         pd_dist = ifelse(policy == 1, (S + E + Ia + Ip + R), 0)) %>%
  summarise(total_death = mean(total_death),
            pd_total = sum(pd_surf) + sum(pd_dist))                                  # add up daily pd over simulation


ef_plot <- vector("list", length = 6)
plot_no <- 0
for(i in seq_along(reduction_intervention)) {
  for(j in seq_along(reduction_iso_mild)) {
     if (reduction_iso_mild[j] > reduction_intervention[i]) next
    # plot efficient frontiers
    plot_no <- plot_no + 1
    ef_plot[[plot_no]] <- summary_df %>%
       filter(
             c_social_dist == reduction_intervention[i],
             c_iso_mild == reduction_iso_mild[j]) %>%
      ggplot(aes(x = pd_total, y = total_death, col = control_policy)) +
      geom_point() +
      scale_x_log10() + 
      theme(legend.position = "bottom", legend.title = element_blank()) +
      facet_wrap(~t_social_dist) +
      labs(title = "Trade-offs between total dead and person-days under social distancing",
           subtitle = paste("Contact reduction for mild and isolated: ", reduction_iso_mild[j], "\n",
                            "Contact reduction during intervention: ", reduction_intervention[i], sep = ""),
           caption = "Person-days calculated for all non-hospitalized and non-symptomatic individuals",
           y = "total dead",
           x = "person days under social distancing (log10 scale)")
    # , caption = "NOTE: Surfing policies cause the outbreak to persist beyond the simulation time,\n 
    # so the total dead and person-days under social distancing are lower bounds for this policy.")
    # facet_grid(c_iso_mild ~ c_social_dist)
  }
}
```

```{r ef_plot, eval = TRUE}
ef_plot[[1]]
ef_plot[[2]]
ef_plot[[3]]
ef_plot[[4]]
ef_plot[[5]]
ef_plot[[6]]
```


```{r eval = FALSE}
susc_plot <- vector("list", length = 6)
plot_no <- 0
for(i in seq_along(reduction_intervention)) {
  for(j in seq_along(reduction_iso_mild)) {
    if (reduction_iso_mild[j] > reduction_intervention[i]) next
    plot_no <- plot_no + 1
    susc_plot[[plot_no]] <-
      sim_df  %>%
      filter(.id %in% sample_ids,
             c_social_dist == reduction_intervention[i],
             c_iso_mild == reduction_iso_mild[j]) %>%
      ggplot() +
      geom_point(aes(x = day, y = 0, col = as.factor(thresh_crossed)), 
                 shape = 22, size = 1, alpha = 0.5) + # show surf policy at bottom
      scale_color_manual(values = c(NA, "black")) +
      geom_line(aes(x = day, y = S, group = .id), col = "purple") +
      #geom_line(aes(x = day, y = H, group = .id), col = "darkgreen") +
      scale_y_log10() +
      facet_grid(t_social_dist ~ control_policy) +
      labs(title = "Epidemic scenarios under alternative management strategies",
           subtitle = paste("Contact reduction for mild and isolated: ", reduction_iso_mild[j], "\n",
                            "Contact reduction during intervention: ", reduction_intervention[i], sep = ""),
           y = "(log10) number of susceptible people",
           x = "days",
           caption = "purple is total infected and non-hospitalized, green is hospitalized cases") +
      theme(legend.position = "n")
    
  }
}
susc_plot[[1]]
susc_plot[[2]]
susc_plot[[3]]
susc_plot[[4]]
susc_plot[[5]]
susc_plot[[6]]
```

# todo

* look at the effect of different thresholds for the surfing policy. Right now, I'm using 20 hospitalized cases to trigger shutdowns which are lifted when hospitalized cases drop to $\leq$ 2.

* Consider including some rules for the "continue social distancing" policy that dictate when social exclusion ends.

* Model death rates as a function of hospital capacity.

For the NSF proposal, should I have economic structure included in the model? I.e., essential workers, work from home, furloughed, laid off.

In a [recent poll of economists about covid-19](http://www.igmchicago.org/surveys/policy-for-the-covid-19-crisis/), this guy, [José Scheinkman](http://www.igmchicago.org/participants/jose-scheinkman/)  stated (without any evidence) that "Optimal strategy involves multiple waves of contact reduction." He might be someone worth reaching out to.

